---
title: "MATH 3070 Lab Project 4"
author: "Harrison Webb"
date: "September 20, 2021"
output:
  html_document:
    toc: TRUE
---
4
*Remember: I expect to see commentary either in the text, in the code with comments created using `#`, or (preferably) both! **Failing to do so may result in lost points!***

## Problem 1 (Verzani problem 2.43)
*The `time` variable in the `nym.2002` data set (**UsingR**) contains the time to finish the 2002 New York City Marathon for a random sample of the finishers.*

1. *What percent ran the race in under 3 hours? (note that the data is in minutes)*

```{r, include=FALSE}
library(UsingR)
```

```{r}
data(nym.2002)
summary(nym.2002)
percentageUnder3 = length(nym.2002$time[nym.2002$time/60 < 3])/length(nym.2002$time) #get values less than 3 from time column
print(percentageUnder3)
```

2. *What is the time cutoff for the top 10%? The top 25%?*

```{r}
# Notice that these are the 90th and 75th percentiles, respectively
# Your code here
quantile(nym.2002$time, c(.75, .90))
```

3. *What time cuts off the bottom 10%?*

```{r}
# Your code here
quantile(nym.2002$time, .1)
```

## Problem 2 (Verzani problem 4.1)
*The data set `UScereal` (**MASS**) contains data on cereals sold in the United States in 1993. For this data set, answer the following questions using R (i.e. MUST answer questions using a code. Do not count by hand):*

1. *How many rows does the data frame have? Columns?*
```{r, inclued=FALSE}
library(MASS)
```

```{r}
# Your code here
data("UScereal")
summary(UScereal)

#rows:
length(UScereal[,1])

#columns
length(UScereal[1,])
```

2. *How many different manufacturers are included?*

```{r}
# Your code here
length(unique(UScereal$mfr)) 
```

3. *How many vitamin categories are included?*

```{r}
# Your code here
length(unique(UScereal$vitamins))
```

4. *How many cereals have a sugar level above 10?*

```{r}
# Your code here
length(UScereal$sugars[UScereal$sugars > 10])
```

5. *What is the mean calorie value for cereals with more than 5 grams of fat? Less than or equal to 5?*

```{r}
# Your code here
#mean calories for cereals with >5 grams of fat
mean(UScereal$calories[UScereal$fat > 5])

#mean calores for cereals with <=5 grams of fat
mean(UScereal$calories[UScereal$fat <= 5])

```

6. *What is the mean calorie value for cereals on the middle shelf (2)?*

```{r}
# Your code here

mean(UScereal$calories[UScereal$shelf == 2])
```

## Problem 3
*Create a data frame containing the data in the following table:*

| First    | Last      | Age |
|----------|-----------|-----|
| Marcus   | Holstein  | 23  |
| Samuel   | Adams     | 56  |
| Gus      | McPherson | 43  |
| Margaret | Olsen     | 41  |
| Zim      | Newbold   | 95  |

```{r}
# Your code here
frame = data.frame("First" = c("Marcus", "Samuel", "Gus", "Margaret", "Zim"),
                   "Last" = c("Holstein", "Adams", "McPherson", "Olsen", "Newbold"),
                   "Age" = c(23, 56, 43, 41, 95))
frame
```
