---
title: "MATH 3070 Lab Project 5"
author: "Harrison Webb"
date: "September 29, 2021"
output:
  html_document:
    toc: TRUE
---

*Remember: I expect to see commentary either in the text, in the code with comments created using `#`, or (preferably) both! **Failing to do so may result in lost points!***


## Problem 1 (Verzani problem 4.7)
*Use the `subset()` function to return a data frame made from the `Cars93` (**MASS**) data frame consisting only of `non-USA` cars in origin, with 4 cylinders and a maximum price of \$15,000 or less.*
```{r, include=FALSE}
library(MASS)
```

```{r, tidy=TRUE,error=TRUE}
# Your code here
data("Cars93")
#summary(Cars93)
subset(Cars93, select = c(Make, Type, Max.Price), subset = Origin == "non-USA" & Cylinders == 4 & Max.Price <= 15)
```

## Problem 2 (Verzani problem 4.10)
*Find the standard deviation of each variable in the `mtcars` data set.*

```{r, tidy=TRUE, error=TRUE}
# Your code here
data(mtcars)
sapply(mtcars, sd) #use sapply to apply standard deviation function to each column in the data set
```

## Problem 3 (Verzani problem 4.11)
*Find the standard deviation for each numeric variable in `Cars93` (**MASS**).*

```{r, tidy=TRUE, error=TRUE}
# Your code here
data("Cars93")
#summary(Cars93)
numericCols = Cars93[sapply(Cars93, is.numeric)] #get columns representing numeric variables
#summary(numericCols)
sapply(numericCols, sd) #apply sd fuction to all columns
```

## Problem 4
*You can find data describing various statistics on 2012 Olympic participating nations such as population and medal counts in the comma-separated value file (olympic-medals2012.csv). Load this dataset into R in a data frame and do the following:*

1. *Create a new data frame containing only the country's ISO code (a three-letter code for the country; USA is the United States of America, and DZA Algeria), name, population, and counts for gold, silver, and bronze. This data frame should be called `olympic2012`.*

```{r, tidy=TRUE, error=TRUE}
# Your code here
readfile = read.csv("olympic-medals2012.csv") #read csv file
olympic2012 = subset(readfile, select = c(ISO, Country.name, pop.2010, Gold, Silver, Bronze)) #select specific columns
olympic2012

```

2. *What proportion of countries earned no medals? Answer this using `olympic2012`.*

```{r, tidy=TRUE, error=TRUE}
# Your code here
numCountries = length(olympic2012[,1])
noMedals = subset(olympic2012, subset = Gold + Silver + Bronze == 0)
numNoMedals = length(noMedals[,1])

numNoMedals / numCountries #proportion of countries that earned no medals
```

