---
title: "MATH 3070 Lab Project 2"
author: "Harrison Webb"
date: "September 2, 2021"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

*Remember: I expect to see commentary either in the text, in the code with comments created using `#`, or (preferably) both! **Failing to do so may result in lost points!***

## Problem 1 (Verzani problem 1.7)
*The `rivers` (**UsingR**) or any other data set is available after loading the package **UsingR**. Load the package, and inspect the data set. Scan the values to find the largest one.*

```{r, include=FALSE}
library("UsingR")
```
```{r}
data(rivers) #load dataset
max(rivers) #use max function to find largest value
```


## Problem 2 (Verzani problem 2.4)
*Create the following sequences, using `:`, `seq()`, or `rep()` as appropriate:*

1. `"a" "a" "a" "a" "a" "a"`

```{r}
rep("a", 6) #repeat "a" 6 times
```

2. `1 3 ... 99` *(the odd numbers)*

```{r}
seq(1, 99, by=2) #start at 1, go to 99, increment by 2 to list odd numbers
```

3. `1 1 1 2 2 2 3 3 3`

```{r}
rep(c(1,2,3), each=3) #repeat each element in (1,2,3) 3 times
```

## Problem 3 (Verzani problem 2.3)
*Let our small data set be* `2 5 4 10 8`.

1. *Enter this data into a data vector `x`.*

```{r}
x = c(2,5,4,10,8)
```

2. *Find the square of each number.*

```{r}
x^2
```

3. *Subtract 6 from each number.*

```{r}
x - 6
```

4. *Subtract 9 from each number and then square the answer.*

```{r}
(x-9)^2
```